---
title: "R Exercise 6: Frequency data (Ch 8)"
author: "Dr. Z"
date: "2026"
output:
  html_document:
    highlight: textmate
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r include = F}
## Textbook followed: The Analysis of Biological Data (3rd Edition) by Whitlock & Schluter
## Chapter 8: Fitting probability models to frequency data

## Activities adapted from the following sources:
## 1) Labs using R made for The Analysis of Biological Data (3rd Edition) by Whitlock and Schluter: https://whitlockschluter3e.zoology.ubc.ca/RLabs/index.html
```

Let's apply what we've learned in Chapter 8 to _R_.
<br>  

__Objectives__

- Calculate the chi-squared test statistic.    
- Be able to conduct a chi-squared test when you: (1) know chi-squared, and (2) when you don't know chi-squared.
- Estimate the µ frequency of Poisson data.  
- Calculate the expected probability and frequency of Poisson data.    
- Be able to conduct a chi-squared hypothesis test under the Poisson model.  

__Shortcuts__:  

| R Command | Notes |
| --- | --- |
| table(df) | creates a frequency table of your dataframe; df is the name of your dataframe; if your dataframe has more than one column, you need specify each column using df$var |
| chisq.test(x, p, rescale.p = T) | runs a chi-squared test; x is your observed data, p is your expected probabilities or values, and rescale.p = T rescales probabilities if they do not equal 1. Do not use for Poisson p-value. | 
| pchisq(q, df, lower.tail = F) | use if you know your chi-squared value AND/OR for the correct p-value for Poisson distributions; q is your chi-squared test statistic, df is your degrees of freedom, and lower.tail = F specifies it is a one-tailed test | 
| dpois(x, lambda) | calculates expected probabilities for Poisson distributions; x is a vector of category values (e.g. 0:20), and lambda is the mean expected frequency (µ) | 
<br>
  
__Reminder__: Save your script for practice at home! :)
<br>  

__Shortcuts__:  

| Symbol or Command | Keyboard Shortcut |
| --- | --- |
| <- | Alt + - |
| # | Shift + 3 | 
| Run one line in script | Ctrl + Enter | 
| Run entire script | Ctrl + Shift + Enter | 
| Open new script | Ctrl + Shift + N  | 
<br>  


__Exercise 5__  
1\. Open _RStudio_ and prepare a new script.

a) Open a new script. All of your code for today's exercises, and your notes and comments will go in this script. Write your filename, title, author, date, and description of the script. 
```{r}
# Filename: (what you will save the script as)
# Title: (give script a title)
# Author: (write your full name here)
# Date: Month Day Year (write the actual date here)

# Description: (describe what this script is for)
```
<br>  

2\. We will learn how to conduct a chi-squared test under the proportional model.  
<br>
__Pt I: Conducting a chi-squared test__  
There are two ways to conduct a chi-squared test under the proportional model in _RStudio_. We will review both methods.

__a\)__ __Method I: You do NOT know X^2__  
If you do not know or have yet to calculate X^2, you can use the function: __chisq.test( )__  

You only need to enter in three pieces of information:  

1. __x__ - your observed values (df$var or vector)  
2. __p__ - your expected proportions (df$var or vector)  
3. __rescale.p = T__ - if your proportions do NOT equal 1, R needs to rescale them to 1 for the test to work  

__b\)__ __Method II: You already know X^2__  
If you already know or calculated X^2 by hand, you can use the function: __pchisq( )__  

You only need to enter in three pieces of information:  

1. __q__ - your observed values (df$var or vector)  
2. __df__ - your expected proportions (df$var or vector)  
3. __lower.tail = T__ - specifies a one-taile test  

___Note___: Chi-squared tests are always one-tailed!  
<br>
__Pt II: Method I with M and Ms__  
M and M's are drawn from a bag and the colour of each M and M is recorded.  

a. Read the file into _R_ from the GitHub using the following URL: https://github.com/lczawadzki/biostats/raw/main/data/MandMlist.csv.
```{r echo = T, results = 'hide'}
     MMlist <- read.csv("https://github.com/lczawadzki/biostats/raw/main/data/MandMlist.csv")
```
b. This dataframe lists the colours, but doesn't tell us the frequency directly. To be able to view the frequency of each colour, we need to sum the data into a table using: __table(df$var)__. Type the following below. Now we can see the frequency of colours in this bag of M and Ms.
```{r echo = T}
     MMtable <- table(MMlist$color)
     MMtable
```

c. Our observed values are in the table. We can assign these to the variable _obs_ using the function __c()__, which will "contain" all of the observed values.
```{r echo = T}
     obs <- c(10,5,12,11,6,11)
     obs
```
d. The M&M company states that 24% are blue, 14% brown, 16% green, 20% orange, 13% red, and 13% yellow. These are our expected proportions. We can assign these to the variable _exp_ using the function __c()__.
```{r echo = T}
     exp <- c(0.24,0.14,0.16,0.20,0.13,0.13)
     exp
```
e. If you are asked for the expected frequency of each M and M, you will multiply the expected proportions by the TOTAL number of MandMs in our sample.
```{r echo = T, results = 'hide'}
     exp*sum(MMtable)
```
f. Now that we have our observed values and expected proportions, we can run a chi-squared test. Remember, the function is __chisq.test__. All you need are your observed values (x), expected proportions (p), and to specify rescale.p = T.
```{r echo = T, results = 'hide'}
     chisq.test(obs, p = exp, rescale.p = T)
```
<br>  
__Understanding your output__  
Let's read through the output. 
The output tells you a number of things about your data:  

1. Line 1 tells you where your observed values are coming from.
2. Line 2 states __X-squared__ which is your chi-squared test statistic, __df__ which is your degrees of freedom, and your ___p___ __-value__. Remember, the _P_-value is the probability that a sample from the null model would be as extreme, or more extreme, than your sample. That is, it is the probability of obtaining your data under the null hypothesis or expectation.  

So, from our output, we know that:  

1. Our data comes from the variable _obs_.
2. Our __X-squared test statisic equals 5.017__, our __df (degrees of freedom) equals 5__, and our ___P___ __-value = 0.4138__.  

We can conclude: "p > 0.05, therefore our results are statistically insignificant, and we fail to reject the H0. Our data match our null expectations, and are likely to occur due to chance. We conclude that the frequency of colours of M and Ms in our bag matches the expectations of the M and M company."  
<br>

__Pt III: Method II with M and Ms__  

If you already calculated X^2 in R or by hand, you can use the function __pchisq__. All you need is your chi-squared test statistic (q), your degrees of freedom (df), and to specify that the lower tail is FALSE.
```{r echo = T, results = 'hide'}
     pchisq(5.017, df = 5, lower.tail = F)
```
__Understanding your output__  
The output of __pchisq__ is one number. This is your ___P___ __-value__. You will end up with the __same__ _P_-value as you did with the output of __chisq.test__, since the df for both is the same.

So, from our output, we know that our ___P___ __-value = 0.4138__.  

We can conclude: "p > 0.05, therefore our results are statistically insignificant, and we fail to reject the H0. Our data match our null expectations, and are likely to occur due to chance. We conclude that the frequency of colours of M and Ms in our bag matches the expectations of the M and M company."  
<br>  
3\. We will learn how to conduct a chi-squared test under the Poisson model.  

It is thought that the World Cup decides the best soccer team in the world. But how much do skill differences determine the outcome? We will look at data on number of goals scored by one or two teams (randomly chosen) in every game of the knockout round of the World Cup from 1986 through 2010 to determine if the number of goals scored is random (follows a Poisson distribution).

a. Read the file into _R_ from the GitHub using the following URL: https://github.com/lczawadzki/biostats/raw/main/data/WorldCup.csv.
```{r echo = T, results = 'hide'}
     worldcup <- read.csv("https://github.com/lczawadzki/biostats/raw/main/data/WorldCup.csv")
     head(worldcup)
```
b. This dataframe lists the number of goals, but doesn't tell us the frequency directly. To be able to view the frequency of each number of goals in a knockout round, we need to sum the data into a table using: __table(df$var)__. Type the following below.
```{r echo = T, results = 'hide'}
     cuptable <- table(worldcup$score)
     cuptable
```
c. Next, we find our mean expected frequency (µ or lambda). Even though we have frequency data, since each observation is listed in a single column, __mean()__ will calculate the mean correctly.
```{r echo = T, results = 'hide'}
     mean(worldcup$score) #this is µ or lambda
```
c. We need to calculate our expected probabilities (these are how we specify the null expectation). We use the function __dpois()__, and specify arguments __x__ which is a vector of the category values, and __lambda__ your expected mean value. We will assign this to the variable _exp_, so that we can plug this directly into our chi-squared function.
```{r echo = T, results = 'hide'}
     exp <- dpois(x = 0:5, lambda = 1.22)
     exp
```
d. Now that we have our observed and expected proportions, we can run a chi-squared test to get our test statistic. Remember, the function is __chisq.test__. All you need are your observed values (x), expected proportions (p), and to specify rescale.p = T. 
```{r echo = T, results = 'hide'}
     chisq.test(obs, p = exp, rescale.p = T)
```
Notice an error has occurred! And p = NA! This error means we have violated an assumption of the chi-squared test. Some categories have expected values that are too small. How can you tell what those expected values are?
```{r echo = T, results = 'hide'}
     expfreq <- nrow(worldcup)*exp
     expfreq
```
We see that one category has an expected value that is less than 1 (5 goals). We need to combine some categories together to meet the assumptions of the chi-squared test. We can combine categories for 4 and 5, so the expected value is more than 1.    

e. Now that we know our categories will be 0, 1, 2, 3, and 4+, we need to create a new vector for our observed data, which we can assign to the variable _obs_. 
```{r echo = T, results = 'hide'}
     cuptable #check original observed values from table
     4 + 1 #find the sum of category 4 and category 5
     obs <- c(32, 44, 21, 10, 5) #observed values
```
f. We also need to create a new vector for our expected proportions, which we can assign to the variable _exp_ We need the expected values to calculate chi-squared later.
```{r echo = T, results = 'hide'}
     exp #check original expected proportions
     0.027251400 + 0.006649342 #find the sum of category 4 and category 5
     exp <- c(0.295230167, 0.360180804, 0.219710290, 0.089348851, 0.03390074) #expected values
     exp
```
g. Now that we have our new observed and expected values that meet assumptions, we can run a chi-squared test to get our test statistic. 
```{r echo = T, results = 'hide'}
     chisq.test(obs, p = exp, rescale.p = T)
```
We can ignore the error this time because we know our expected values meet assumptions. It is just warning us that we have an expected value less than 5.

g. Then we will use __pchisq__ to get our _P_-value. The reason we do NOT use the _P_-value from __chisq.test__ is because it does NOT use the correct degrees of freedom for a Poisson model. Remember, that df = number of categories - 1 - number of parameters. Since we calculated one parameter (µ) to make our estimations, the df for a Poisson model is df = number of categories - 1 - 1. For our data, this is df = 5 - 1 - 1 = 3.
```{r echo = T, results = 'hide'}
     pchisq(1.2741, df = 3, lower.tail = F)
```
So, from both outputs, we know that:  

1. Our data comes from the variable _obs_.
2. Our __X-squared test statisic equals 1.2741__.
3. Our ___P-value___ with the correct df, equals ___0.735__.

We can conclude: "p > 0.05, therefore our results are statistically insignificant, and we fail to reject the H0. We conclude that the number of goals scored in the knockout round of the World Cup follows a Poisson distribution.

Is this supported by the variance? Recall, if the variance = mean then data follow a Poisson distribution.
```{r echo = T, results = 'hide'}
     var(worldcup$score)
```
The variance = 1.256, which is close to the mean of 1.223. Additionally, the ratio of variance to mean = 1.02, which is very close to 1. We thus expect the data to follow a Poisson distribution.  
<br>
**Now it's time to practice on your own.**  

1. Let's look at a sample of birthdays over different days of the week.  

| Day of week | # of birthdays |  
| --- | --- |
| Monday | 33 |
| Tuesday | 41 |
| Wednesday | 63 |
| Thursday | 63 |
| Friday | 47 |
| Saturday | 56 |
| Sunday | 47 |
| Total	| 350 |  

    a\. Enter the observed data (counts of birthdays across weekdays)   
    
    b\. Assume we expect equal numbers of birthdays on all days of the week. What is your null hypothesis?  
    
    c\. Conduct a chi-squared test to test this hypothesis. Note, if we expect equal numbers of birthdays on all days of the week, we do not need to enter "p" in the chi-squared test.  
    
    d\. Do you reject or fail to reject your H0?  
      
    ```{r include = F}
#Answer:
obs <- c(33, 41, 63, 63, 47, 56, 47)
#H0: The numbers of birthdays will be the same on every day of the week.
chisq.test(obs)
#p < 0.05, reject H0
    ```