---
title: "R Exercise 3: Displaying and describing data (Ch 2, 3)"
author: "Dr. Z"
date: "2026"
output:
  html_document:
    highlight: textmate
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r include = F}
## Chapters 2 & 3: Displaying and describing data

## Activities drawn from the following sources:
## 1) Labs using R made for The Analysis of Biological Data (3rd Edition) by Whitlock and Schluter: https://whitlockschluter3e.zoology.ubc.ca/RLabs/index.html
## 2) Statistics for Linguists by Winter
## 3) Getting Started with R (2nd Edition) by Beckerman, Childs, and Petchey
```

Let's apply what we've learned in Chapters 2 and 3 to _R_.
<br>  
__Objectives__  

- Review reading in data.
- Be able to create basic graphs: histograms and box plots.
- Know how to interpret histograms and boxplots.
- Be able to calculate summary statistics for location (central tendency).
- Be able to calculate summary statistics for spread.
<br>
  
__Reminder__: Save your script for practice at home! :)
<br>  

__Shortcuts__:  

| Symbol or Command | Keyboard Shortcut |
| --- | --- |
| <- | Alt + - |
| # | Shift + 3 | 
| Run one line in script | Ctrl + Enter | 
| Run entire script | Ctrl + Shift + Enter | 
| Open new script | Ctrl + Shift + N  | 
<br>  


__Exercise 3__  
1\. Open _RStudio_ and prepare a new script.

a) Open a new script. All of your code for today's exercises, and your notes and comments will go in this script. Write your filename, title, author, date, and description of the script. 
```{r}
# Filename: (what you will save the script as)
# Title: (give script a title)
# Author: (write your full name here)
# Date: Month Day Year (write the actual date here)

# Description: (describe what this script is for)
```
<br>  

2\. Let's work with the file "student-height.csv" in the GitHub.

__Pt I: Reading in the file__  
Read the file into _R_, either directly from the GitHub using the following URL: https://github.com/lczawadzki/biostats/raw/main/data/student-height.csv, or by downloading the data, and using the pathname/working directory. Remember to assign it to a name.  
```{r echo = T, results = 'hide'}
     studht <- read.csv("https://github.com/lczawadzki/biostats/raw/main/data/student-height.csv", header = TRUE)
```
<br>  
__Pt II: Displaying data (view format, make histogram)__  
a. Look at the structure of the dataframe using the function __str( )__. This tells you information about the dataframe, including how many observations, how many variables, and what those variables are.
```{r echo = T}
     str(studht)
```
<br>
b\. Use the function __head( )__ to look at the first few lines of the dataframe.
```{r echo = T, results = 'hide'}
     head(studht)
```
<br>
c\. Our data consists of one numerical variable. To plot one numerical variable, we use a __histogram__. To plot histograms in _R_, we can use the function __hist( )__. However, we cannot just type in the dataframe as the argument, we need to specify both the dataframe and the variable, using the notation: _dataframe$variable_.
```{r echo = T, fig.keep = "none"}
     hist(studht$height)
```
<br>
d\. We can change the labels by adding the argument _xlab = ""_ for the x-axis label, and _main = ""_ for the title of the graph. Test this out.
```{r echo = T, fig.keep = "none"}
     hist(studht$height, xlab = "Student Height (in)", main = "Frequency of Student Height")
```
<br>
__e\. Now, look at your histogram__:  
- What is the __shape__ of our distribution (symmetric, asymmetric)?  
- How many __peaks__ do we have (this is the mode)?  
<br>  
__Pt III: Describing data (summary statistics)__  
a. Let's calculate our summary statistics, using the following functions. __Remember__, we need to specify which variable we are looking at using the notation: _dataframe$variable_.
```{r echo = T, results = 'hide'}
     mean(studht$height)
     median(studht$height)
     sd(studht$height)  #standard deviation
     var(studht$height) #variance
     IQR(studht$height) #interquartile range
```
__Note:__ we do ___not___ have a function for the __mode__. To find the mode, we can always look at the __peak__ of our histogram!
<br>  
b\. We also do ___not___ have a function for the __coefficient of variation__. We will need to calculate this directly, using our knowledge of the formula, and the functions for standard deviation (__sd__) and mean (__mean__). The formula is: CV = (sd/mean)*100%. __Remember__, we need to specify which variable we are looking at using the notation: _dataframe$variable_.
```{r echo = T, results = 'hide'}
     (sd(studht$height)/mean(studht$height))*100
```
<br>
c\. A faster way to get all of these statistics in one go is to use the __summary( )__ function. This will give you the minimum value, first quartile, median, mean, third quartile, and maximum value. You will still need to calculate the standard deviation (__sd__),variance (__var__), and __interquartile range (__IQR__) separately if you use __summary( )__!
```{r echo = T}
     summary(studht)
```
<br>
d\. We can also make a box plot of our data, to examine the median, first quartile, third quartile, and interquartile ranges. The function for a box plot is __boxplot( )__. You can add labels in the same way you did with your histogram, except this time we'll want to label the y-axis using _ylab = ""_. The argument for boxplot is just _dataframe_.
```{r echo = T, fig.keep = "none"}
     boxplot(studht, ylab = "Student Height (in)", main = "Box Plot of Student Heights")
```
__e\. Now, examine your boxplot__:  
- What does the black line represent?  
- What do the edges of your box represent?  
- What does the range from one end of the box to the other represent?  
- What do the whiskers show?  
<br>  
**Let's practice with another file.** 
  
1. Read in the file "bird-richness.csv" from the GitHub using the following URL: https://github.com/lczawadzki/biostats/raw/main/data/bird-richness.csv (or by downloading the data). Assign it the name "birds". 
```{r include = F}
#Answer:
birds <- read.csv("https://github.com/lczawadzki/biostats/raw/main/data/bird-richness.csv", header = T)
```
    a\. View the data using str() and head().  
    
    b\. Suppose we are interested in the number of birds found at a particular location (the Richness column). We would make a histogram. Plot a histogram, using the notation: _dataframe$variable_.   
    
    c\.Plot a boxplot of species richness using the notation: _dataframe$variable_.
    
    d\. What if we wanted to do a comparison of bird species richnesses in different habitats? [Possible habitats in this file are “forest” and “grassland”.] In your boxplot, you will need to use the notation _y~x_, where _y_ is the variable on the _y-axis_, and _x_ the variable on the _x-axis_. Plot the following.  
```{r echo = T, fig.keep = "none"}
     boxplot(birds$Richness ~ birds$habitat)
```
      
  e\. What pattern do you see? What is the effect of habitat on species richness? How do the medians vary? The IQR? 

```{r include = F}
#Answer:
str(birds)
head(birds)
hist(birds$Richness)
boxplot(birds$Richness)
boxplot(birds$Richness~birds$habitat)
```
```{r include = F}
     
#NEED A DIFFERENT DATASET FOR PRACTICE, ONE VARIABLE SO THEY CAN DO MEAN ETC. MAYBE SNAKE EXAMPLE? ADD LATER. MAYBE FOR NEXT EXERCISE CH 3 AND 4. NO TIME.
```